Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-12-14 WholePlotTrajectories.tex
review

Disturbances change both biotic and abiotic environment, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and of the interactions among species such as
competition \citep{Chesson2000}. Community response has been largely
studied through forest structural parameters such as aboveground
biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Thereafter, models based on the
observed trajectories after disturbance gave important insights into the
recovery of ecosystem processes and services \citep{Herault2018}.
Regarding community diversity and composition, however, post-disturbance
trajectories are not as thoroughly understood. It is recognized that
determined succession, mainly driven by the recruitment of pioneer and
heliophilous species, shape post-disturbance trajectories
\citep{Carreno-Rocabado2012, Poorter2016}. Community trajectories would
then rely on deterministic processes and would be predictable according
to the disturbance intensity and to the characteristics of the
pre-disturbance community. Yet, observed trajectories in diversity and
composition often deviate from predictions. Besides, although the
deterministic ecological processes at stake are identified, their
relative importance compared to stochasticity remains debated
\citep{Norden2015}. This debate translates into the validation and the
extent of the intermediate disturbance hypothesis (IDH). The IDH assumes
the predominance of deterministic processes, and predicts a maximum
species diversity after moderate disturbance intensity
\citep{Molino2001, Kariuki2006a}. Validations of the IDH remain scarce
in the long-term and often rely on the analysis of taxonomic richness
\citep{Molino2001}. Taxonomic richness alone, however, gives limited or
misleading information on forests recovery and functioning
\citep{Chaudhary2016}. More ecological-meaningful analysis would couple
richness with species evenness, which reveals the changes in the species
abundance distribution, and (ii)species composition, which is crucial
for conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore,
a functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines community resilience after
disturbance \citep{Elmqvist2003, Diaz2005}.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

Community taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years.

The taxonomic differences among local communities were already apparent
before disturbance, as revealed by the distinct starting points on the
NMDS axis 2. These differences were maintained throughout recovery
trajectories.\\
More than commonly thought, post-disturbance trajectories depended on
community initial composition, which partly determined the pool of
recruited species and constrained the trajectories towards the initial
composition. The community taxonomy proved highly resilient, as it
absorbed the disturbance and maintained commnity initial composition
characteristics \citep{Folke2006}. This high resilience suggested that
species not belonging to the pre-disturbance community were rarely
recruited because of the commonness of dispersal limitation among
tropical tree species \citep{Svenning2005}.

Conversely, disturbed communities followed functional trajectories that
are highly similar in terms of functional composition. As the
composition of pre-disturbance surviving trees is representative of the
initial community \citep{Herault2018}, changes in functional composition
relied upon the recruitment of species or functional types that were
infrequent or absent before disturbance. Competitive pioneers became
dominant in filling the environmental niches of high availability of
light, space and nutrients vacated by the disturbance. The recruitment
of pioneers changed community functional composition in the same way for
all disturbance intensity towards more resource-acquisitive strategies,
moving community functional composition right along the first axis in
Fig. \ref{fig:NMDSplans} \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter long-lived, more resistant and shade-tolerant species
excluded the first established pioneers and started the recovery of
pre-disturbance functional composition, moving similarly community
functional composition left along the first axis and upward along the
second axis in Fig. \ref{fig:NMDSplans}.

These trajectories provided empirical support to the hypothesis that
community assembly is both deterministic and historically convergent at
different levels of community organization. Deterministic, trait-based
processes drove community convergence in functional composition, while
at the same time dispersal limitation maintained their divergence in
taxonomic composition \citep{Fukami2005}.

@@

Community taxonomic richness increased after disturbance until an
intensity threshold (20-25\% AGB loss) above which the taxonomic
richness decreased. The increase of taxonomic evenness after disturbance
also depended on the disturbance intensity, although the correlation was
weaker than for the taxonomic richness as already observed in the Guiana
Shield \citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. In accordance with the IDH, then, community
taxonomic trajectories depended on the disturbance intensity. They
markedly changed above an intensity threshold, with a decrease of the
richness and a persistent increase of the evenness. Disturbance
intensity determined the balance in the community between surviving
trees from pre-disturbance communities and trees recruited afterward.
The pool of true pioneer species specifically recruited after
disturbance is, in the Guiana Shield, restricted to a few common genera
(e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)\citep{Guitet2018}. Below
the intensity threshold the surviving community trees remained numerous
enough to maintain the pre-disturbance high taxonomic richness while the
pioneers recruited, infrequent or absent before disturbance, increased
both community taxonomic richness and evenness. Beyond the intensity
threshold, disturbance decreased the taxonomic richness of surviving
trees, and this decrease was not offset by the recruitment of pioneers.
The overall community taxonomic richness therefore decreased according
to the disturbance intensity \citep{Molino2001}. An intermediate
disturbance was then detected, defined by an intensity for which
post-disturbance trajectories markedly changed. Such disturbance
increased the availability of resources and created opportunities for
pioneers, without implying too important loss of shade-tolerant species
or preventing their maintenance in the community \citep{Bongers2009}.
For community taxonomic evenness, the disturbance impact was similar but
milder. Because the evenness is less sensitive to the loss of rare
species, below the intensity threshold the trajectories of taxonomic
evenness rather represented the increasing dominance of pioneers that
balancing the usual hyper-dominance of a few species. Beyond the
intensity threshold, however, pioneers became in turn highly dominant
and decreased the overall evenness \citep{Baraloto2012a}.

Conversely the IDH was disproved regarding the disturbance impact on
community functional richness and evenness. Irrespective of the
disturbance intensity, both community functional richness and evenness
increased following the recruitment of pioneers, functionally highly
different from the composition of pre-disturbance community.

Along time, taxonomic richness trajectories of all disturbed communities
similarly dropped at first place, following the species loss due to
disturbance, and then displayed a species gain depending on the
disturbance intensity. Up to an intensity threshold, the species gain
was all the more significant that the disturbance intensity increased,
with the establishment of long-lived pioneers enhancing community
taxonomic richness and evenness in the long term. These long-lived
pioneers, functionally quite different from the functional composition,
entailed as well a progressive and long-lasting increase of the
functional richness and evenness \citep{Denslow1980, Molino2001}. Beyond
an intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness and evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance \citep{Walker2009}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, significantly decreased the functional redundancy
within the pre-disturbance functional space. The redundancy decrease was
not compensated for in the first place as the first recruited pioneers
were functionally different from the pre-disturbance functional
composition. Progressively though, first established species were
replaced by more competitive long-lived pioneers or late-successional
species resembling more the pre-disturbance functional composition and
restoring the functional redundancy. This replacement was stochastic and
followed the lottery recruitment rules, implying a recruitment eased for
the first recruited species but then increasingly hampered by the
emergence of interspecific competition \citep{Busing2002}. Along time
the recovery of infrequent species was increasingly slow, so that the
time to full recovery of the functional redundancy, in some communities
just initiated after 30 years, was extremely difficult to estimate
\citep{Elmqvist2003, Diaz2005}.

The mid-term impact of disturbance on community functional redundancy
meant a lower resilience of the pre-disturbance communities, with higher
chances to see the persistence of disturbance-specific species at the
expense of late-successional ones \citep{Haddad2008}. Besides, the
mid-term recovery of infrequent species increases the risks to loose
keystone species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Apart from the functional
characteristics considered here, infrequent species might indeed have
unique functions in the ecosystem or be a key for some fauna
\citep{Schleuning2016}.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evenness which were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functional composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pioneers that prevented in the first years after disturbance
the establishment of more diverse long-lived pioneers, recruited
otherwise below the intensity threshold. The resilience of tropical
forests, defined in terms of recovery to pre-disturbance state, proved
tangible but requiring several decades. Still, the disturbance impact on
community redundancy cautioned against the risks of infrequent species
loss and the persistence of disturbance-specific communities
\citep{Herault2018}.

>>>>

Disturbances change both biotic and abiotic environment, through
modifications in the fluxes of light, heat and water
\citep{Goulamoussene2017}, and modifications in the interactions among
species \citep{Chesson2000}. Community response has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016}.
Thereafter, models based on the observed trajectories after disturbance
gave important insights into community recovery and resilience
\citep{Herault2018}. Regarding community diversity and composition,
however, post-disturbance trajectories are not as thoroughly understood.
It is recognized that a determined succession, mainly driven by the
recruitment of pioneer and heliophilous species, shape community
post-disturbance trajectories \citep{Carreno-Rocabado2012, Poorter2016}.
Deterministic processes would then be major processes of community
response to disturbance. This would make post-disturbance trajectories
deterministic, and then predictable according to the disturbance
intensity and to the characteristics of the pre-disturbance community.
Yet, observed trajectories in diversity and composition often deviate
from predictions. Although the deterministic ecological processes at
stake are identified, their interplay with stochastic processes and
their relative importance remains debated \citep{Norden2015}. This
debate translates into the validation and the extent of the intermediate
disturbance hypothesis (IDH). The IDH assumes the predominance of
deterministic processes, and predicts a maximum species diversity after
moderate disturbance intensity \citep{Molino2001, Kariuki2006a}. The IDH
has been scarcely validated in the long-term and often rely on the
analysis of taxonomic richness \citep{Molino2001}. Taxonomic richness
alone, however, gives limited or misleading information on forests
recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with species
evenness, which reveals the changes in the species abundance
distribution, and (ii)species composition, which is crucial for
conservation issues \citep{Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link community diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually,
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines community resilience after
disturbance \citep{Elmqvist2003, Diaz2005}.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1, low & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2, intermediate & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3, high & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

The hump-shaped trajectories of community composition returned towards
pre-disturbance state after 30 years, highlighting community resilience
in terms of taxonomic and functional composition.

Before disturbance, there were already taxonomic differences among local
communities, as revealed by the distinct starting points on the NMDS
axis 2. These initial differences were maintained throughout the
trajectories following disturbance, so community initial state
influenced the trajectories following disturbance more than commonly
thought. The initial composition indeed partly determined the pool of
species recruited, and hence shaped the composition trajectories and
drove them towards the recovery of the initial composition. This
taxonomic recovery, although it was not fully achieved after 30 years,
suggested the resilience of community taxonomic composition and the
maintenance of community initial differences \citep{Folke2006}. Species
not belonging to the pre-disturbance community were then rarely
recruited, probably because of the common dispersal limitations among
tropical tree species \citep{Svenning2005}.

Functional composition trajectories were, conversely, highly similar
among disturbed communities. As the composition of pre-disturbance
surviving trees is representative of the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types that were infrequent or
absent before disturbance. Competitive pioneers became dominant in
filling the environmental niches vacated by the disturbance, with a high
availability of light, space and nutrients. This recruitment of pioneers
similarly changed the functional composition of all communities,
whenever the disturbance intensity. The functional composition went
towards more resource-acquisitive strategies, translated in the
functional plan (Fig. \ref{fig:NMDSplans}) by a displacement on the
right along the first axis \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter, the pioneers first recruited were progressively excluded by
long-lived, more resistant and shade-tolerant species. The commpunity
functional composition then returned towards more resource-conservative
strategies, suggesting the recovery of the initial community composition
and translated in the functional plan by a displacement left along the
first axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

In the light of these taxonomic and functional composition trajectories,
community response to disturbance proved simultaneously deterministic
and stochastic at different levels of community organization. Taxonomic
trajectories were shaped by dispersal limitations maintaining the
stochastic differences among communities due to historical contingency.
Functional trajectories were shaped by trait-based processes resulting
the the convergence of community functional composition
\citep{Fukami2005}.

@@

Community taxonomic richness and evenness trajectories both proved to
depend on the intensity of disturbance and to markedly change above an
intensity threshold, thus supporting the IDH. The disturbance intensity
determined the balance in the community between pre-disturbance
surviving trees, and trees recruited afterward. Below an intensity of
20\%-25\% AGB loss, the trees surviving after disturbance remained
numerous enough to maintain the high taxonomic richness of the
pre-disturbance community. The recruitment of pionners, infrequent or
absent before disturbance, then increased the community taxonomic
richness positively correlated with the disturbance intensity
\citep{Martin2015, Chaudhary2016}. Besides, the pioneers became
increasingly dominant and temporarily balanced the usual hyper-dominance
of a few species \citep{Baraloto2012a}. The taxonomic evenness then
followed hump-shaped trajectoriesas already observed in the Guiana
shield \citep{Baraloto2012a} and in Borenan tropical forests
\citep{Cannon1998}, with first an increase positively correlated to the
disturbance intensity, and then an return towards initial values. Above
the intensity threshold, the taxonomic richness of surviving trees was
too low to be offset by the recruitment of pioneers. In the Guiana
shield indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
trajectories following intense disturbance then decreased, all the more
so that the disturbance intensity was high \citep{Molino2001}. At the
same time, pioneers became persistently dominant and prevented a return
towards the initial evenness and the recovery of hyper-dominant
shade-tolerant species. The intensity threshold then marked two distinct
post-disturbance trajectories, and a change in community response to
disturbance. In accordance with the IDH, an intense disturbance below
the intensity threshold would then maximise community richness and
evenness. Such intermediate disturbance would increase the availability
of resources, creating opportunities for pioneers, while allowing a high
diversity among surviving trees and the maintenance of shade-tolerant
species \citep{Bongers2009}.

Regarding community functional response in contrast, the IDH seemed
disproved. Whenever the disturbance intensity, both community functional
richness and evenness increased, with the recruitment of pioneers that
were functionally highly different from the species of the
pre-disturbance community \citep{Denslow1980, Molino2001}. Above the
intensity threshold however, for the most intense disturbance, community
functional richness and evenness started to decrease after 15 to 20
years. The vacated environmental space was then occupied by short-lived
species which prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and the taxonomic and functional trajectories will catch up
with those observed for intermediate disturbance \citep{Walker2009}.

The IDH then translated into community post-disturbance taxonomic
trajectories, with marked differences below and above an intensity
threshold. It was however not tangible regarding the functional
trajectories, because of their decoupling with taxonomic trajectories.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying a recruitment easy for the first recruited
species but increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The functional redundancy then relied
upon stochastic processes and its recovery, depending on the recruitment
of infrequent species constrained by the increasing competition, was
increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

The impact of disturbance on community functional redundancy meant a
lower resilience of the pre-disturbance communities, with higher chances
to see the persistence of disturbance-specific species at the expense of
late-successional ones \citep{Haddad2008}. Besides, the stochastic
recovery of infrequent species increases the risks to lose keystone
species, with unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005}. Apart from the functional
characteristics considered here, infrequent species might indeed have
unique functions in the ecosystem or be a key for some fauna
\citep{Schleuning2016}.

@@

Our study highlighted the combination of deterministic and stochastic
processes determining the succession of recruited species and shaping
community response to disturbance. The deterministic recruitment of
pioneers shaped functional trajectories convergent in the functional
space, while dispersal limitations shaped stochastic taxonomic
trajectories maintaining community divergence in taxonomic composition.
In accordance with the IDH, community taxonomic diversity was maximise
for intermediate disturbance intensity, before reaching an intensity
threshold beyond which the taxonomic richness decreased and the
taxonomic evenness remained persistently higher. Conversely, functional
trajectories proved decoupled from the taxonomic ones and did not follow
the hypothesis of the IDH as no distinct trajectories emerged with
increasing disturbance intensity. The decoupling between taxonomic and
functional disturbance was explained by a decrease of community
functional redundancy mitigating the functoinal impact of disturbance.
The resilience of tropical forests, defined in terms of recovery to
pre-disturbance state, proved tangible but requiring several decades.
Still, the disturbance impact on community redundancy cautioned against
the risks of infrequent species loss and the persistence of
disturbance-specific communities \citep{Herault2018}.
